---
title: "Getting started with inatpick"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with inatpick}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE
)
```

## Overview

`inatpick` provides a simple workflow for downloading photos and metadata
from [iNaturalist](https://www.inaturalist.org) via its API. The three main
functions are:

| Function | Description |
|---|---|
| `inat_fetch()` | Fetch observations from the API |
| `inat_download()` | Download photos to a local folder |
| `inat_metadata()` | Export observation metadata to CSV |

```{r}
library(inatpick)
```

---

## Example: *Drosera rotundifolia* in the United Kingdom

In this example we download all observations of the round-leaved sundew
(*Drosera rotundifolia*, taxon ID `52658`) recorded in the United Kingdom
(place ID `6857`) by iNaturalist user
[andresrb](https://www.inaturalist.org/people/andresrb).

### Step 1 — Fetch observations

```{r}
obs <- inat_fetch(
  taxon_id   = 52658,      # Drosera rotundifolia
  place_id   = 6857,       # United Kingdom
  user_login = "andresrb"
)

nrow(obs)  # how many observations?
```

### Step 2 — Filter by annotation

To focus only on observations where flowers are visible, use the `annotation`
argument. Run `inat_annotations` to see all available options.

```{r}
# See all annotation labels
inat_annotations

# Fetch only flowering observations
obs_flowering <- inat_fetch(
  taxon_id   = 52658,
  place_id   = 6857,
  user_login = "andresrb",
  annotation = "flowers"
)
```

### Step 3 — Download photos

Download photos to a local folder. The `size` argument controls resolution:
`"square"` (75px), `"small"` (240px), `"medium"` (500px), `"large"` (1024px,
default), or `"original"`.

```{r}
inat_download(
  obs,
  out_dir = "drosera_photos",
  size    = "large"
)
```

Files are named `obs{observation_id}_{photo_id}.jpg` and saved to the
specified folder. Already-downloaded files are skipped by default
(`overwrite = FALSE`).

### Step 4 — Export metadata to CSV

```{r}
inat_metadata(
  obs,
  path = "drosera_rotundifolia_uk.csv"
)
```

The CSV includes observation ID, date, taxon name, observer, place, 
coordinates, and number of photos.

---

## Finding taxon and place IDs

IDs are visible directly in iNaturalist URLs:

- **Taxon ID**: `inaturalist.org/taxa/52658-Drosera-rotundifolia` → `52658`
- **Place ID**: `inaturalist.org/observations?place_id=6857` → `6857`

---

## Photo licenses

Photos downloaded from iNaturalist retain their original licenses as set
by individual observers (CC0, CC-BY, CC-BY-NC, etc.). Use
`licensed = TRUE` in `inat_fetch()` to restrict to CC-licensed photos only.
Always credit the observer and respect the license terms when using images
in research or publication.

```{r}
# Only CC-licensed observations
obs_cc <- inat_fetch(
  taxon_id = 52658,
  place_id = 6857,
  licensed = TRUE
)
```

---

## Further filtering options

```{r}
# Research-grade only, flowering, in July
obs_july <- inat_fetch(
  taxon_id      = 52658,
  place_id      = 6857,
  annotation    = "flowers",
  quality_grade = "research",
  month         = 7
)
```
